---
title: "PCA analyses all populations"
author: "ASH IWIKAU"
date: "2026-01-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Load libraries
```{r message=FALSE}
## load plotting packages
library("tidyverse")
library("ggplot2")
library("ggrepel")
library(plotly)

```


Read in eigenvector/value data generated by PLINK
```{r}
## read in eigenvectors/values
eigenvec = read.delim("all_samples_pca.eigenvec", header = TRUE, sep ="\t")
eigenval = read.delim("all_samples_pca.eigenval", header = FALSE)

## read in metadata
meta = read.delim("../00_metadata/all_locations_no_lowdata_FULL_METADATA.txt", header = TRUE, sep = "\t")
colnames(meta) = c("IID", "MIGRATORYSTATUS", "POP", "ISLANDMAINLAND", "MAINLAND2", "DIADROMOUS")

```



Merge metadata and eigenvector data for easy plotting
```{r}
## merge
fulldata = merge(eigenvec, meta, by = "IID")

## turn to df
fulldata = as.data.frame(fulldata)

```


Calculate percentage of variance explained by each PC for the axes
```{r}
percent_var = (eigenval$V1 / sum(eigenval$V1)) * 100

eigenval <- add_column(eigenval, PC = 1:10, .before = "V1")

plot(eigenval, type = "b",
     xlab = "Principal Component",
     ylab = "Eigenvalue",
     main = "Scree Plot")

ggplot(eigenval, aes(x = PC, y = V1, group = 1)) +
  geom_point(size = 3) +
  geom_line() +
  xlab("Principal Component") +
  ylab("Proportion of Variance Explained") +
  ggtitle("Scree Plot") +
  theme_minimal()

```


### Assign colours to each population for consistency
```{r}
colours = c("Auckland-Islands" = "#fde725", "Breccia-Ck" = "#c8e020", "Chatham-Islands" = "#90d743", "Green-Lake" = "#5ec962", "Lagoon-Saddle-Tarn" = "#35b779", 
            "Lake-Chalice" = "#20a486", "Lake-Marian" = "#21918c", "Lake-Paringa" = "#287c8e", "Lake-Rotoroa" = "#31688e", "Lake-Sylvester" = "#3b528b", 
            "Stewart-Island" = "#443983", "Thirteen-Mile-Ck" = "#481f70", "Twelve-Mile-Ck" = "#440154", "Diadromous" = "#fde725" )

colours = c("Auckland-Islands" = "#F4C40FFF", "Breccia-Ck" = "#FE9B00FF", "Chatham-Islands" = "#D8443CFF", "Green-Lake" = "#9B3441FF", "Lagoon-Saddle-Tarn" = "#DE597CFF", 
            "Lake-Chalice" = "#E87B89FF", "Lake-Marian" = "#E6A2A6FF", "Lake-Paringa" = "#AA7AA1FF", "Lake-Rotoroa" = "#9F5691FF", "Lake-Sylvester" = "#633372FF", 
            "Stewart-Island" = "#1F6E9CFF", "Thirteen-Mile-Ck" = "#2B9B81FF", "Twelve-Mile-Ck" = "#92C051FF", "Diadromous" = "#fde725")


diadromous_colours = c("Diadromous" = "#F4C40FFF", "Non-Diadromous" = "#E87B89FF")
  
island_colours = c("Island" = "#F4C40FFF", "Mainland" = "#E87B89FF")

## lake marian and lake paringa are currently way too close in colour - fix somehow
```

## Plot PCA
### PC1/PC2
```{r}
ggplot(fulldata, aes(x = PC1, y = PC2, color = POP)) +
  geom_point(aes(color = POP, shape = MIGRATORYSTATUS), size = 4, alpha = 0.8) +
  scale_color_manual(values = colours) +
  scale_shape_manual(values = c("Diadromous" = 17, "Non-Diadromous" = 16)) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(percent_var[2], 2), "% variance")) +
  ##geom_text_repel(aes(label = POP), size = 3) +
  theme_minimal()

```

## PC2/PC3
```{r}
ggplot(fulldata, aes(x = PC2, y = PC3, color = POP)) +
  geom_point(size = 3) +
  scale_color_manual(values = colours) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(percent_var[2], 2), "% variance")) +
  ##geom_text_repel(aes(label = POP), size = 3) +
  theme_minimal()

```


## PC3/PC4
```{r}
ggplot(fulldata, aes(x = PC3, y = PC4, color = POP)) +
  geom_point(size = 3) +
  scale_color_manual(values = colours) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(percent_var[2], 2), "% variance")) +
  ##geom_text_repel(aes(label = POP), size = 3) +
  theme_minimal()

```


## PC4/PC5
```{r}
ggplot(fulldata, aes(x = PC4, y = PC5, color = POP)) +
  geom_point(size = 3) +
  scale_color_manual(values = colours) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(percent_var[2], 2), "% variance")) +
  ##geom_text_repel(aes(label = POP), size = 3) +
  theme_minimal()

```

## Thoughts:

Do see the thing where each PC is sort of splitting off a different population from the group altho the islands are still p far out in all

Remove islands and rerun to see in-group relationships btwn mainland pops?


## No Chathams

## Re-calculated eigenvectors for VCF - islands

Read in eigenvector/value data generated by PLINK
```{r}
## read in eigenvectors/values
nochath_eigenvec = read.delim("nochathams_pca.eigenvec", header = TRUE, sep ="\t")
nochath_eigenval = read.delim("nochathams_pca.eigenval", header = FALSE)

```


Merge metadata and eigenvector data for easy plotting
```{r}
## merge
nochath = merge(nochath_eigenvec, meta, by = "IID")

## turn to df
nochath = as.data.frame(nochath)

```


Calculate percentage of variance explained by each PC for the axes
```{r}
nochath_percent_var = (nochath_eigenval$V1 / sum(nochath_eigenval$V1)) * 100

```

## Plot PCA
### PC1/PC2
```{r}
ggplot(nochath, aes(x = PC1, y = PC2, color = POP)) +
  geom_point(size = 3) +
  scale_color_manual(values = colours) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(nochath_percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(nochath_percent_var[2], 2), "% variance")) +
  ##geom_text_repel(aes(label = POP), size = 3) +
  theme_minimal()

```

## No islands

## Re-calculated eigenvectors for VCF - islands

Read in eigenvector/value data generated by PLINK
```{r}
## read in eigenvectors/values
noisland_eigenvec = read.delim("no_islands_pca.eigenvec", header = TRUE, sep ="\t")
noisland_eigenval = read.delim("no_islands_pca.eigenval", header = FALSE)

```


Merge metadata and eigenvector data for easy plotting
```{r}
## merge
noislands = merge(noisland_eigenvec, meta, by = "IID")

## turn to df
noislands = as.data.frame(noislands)

```


Calculate percentage of variance explained by each PC for the axes
```{r}
noisland_percent_var = (noisland_eigenval$V1 / sum(noisland_eigenval$V1)) * 100

```

## Plot PCA
### PC1/PC2
```{r}
ggplot(noislands, aes(x = PC1, y = PC2, color = POP)) +
  geom_point(size = 3) +
  scale_color_manual(values = colours) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(noisland_percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(noisland_percent_var[2], 2), "% variance")) +
  ##geom_text_repel(aes(label = POP), size = 3) +
  theme_minimal()

```

## Thoughts

Now LST is separated and everything else is still kind of horrible

Remove LST and plot again yay

## No islands + LST

Read in eigenvector/value data generated by PLINK
```{r}
## read in eigenvectors/values
LST_eigenvec = read.delim("no_lagoon_no_islands_pca.eigenvec", header = TRUE, sep ="\t")
LST_eigenval = read.delim("no_lagoon_no_islands_pca.eigenval", header = FALSE)

```


Merge metadata and eigenvector data for easy plotting
```{r}
## merge
lst = merge(LST_eigenvec, meta, by = "IID")

## turn to df
lst = as.data.frame(lst)

```


Calculate percentage of variance explained by each PC for the axes
```{r}
lst_percent_var = (LST_eigenval$V1 / sum(LST_eigenval$V1)) * 100

```

## Plot PCA
### PC1/PC2
```{r}
ggplot(lst, aes(x = PC1, y = PC2, color = POP)) +
  geom_point(size = 3,alpha = 0.9) +
  scale_color_manual(values = colours) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(lst_percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(lst_percent_var[2], 2), "% variance")) +
  ##geom_text_repel(aes(label = POP), size = 3) +
  theme_minimal()

```

### Diadromous metapop

```{r}
ggplot(lst, aes(x = PC1, y = PC2, color = DIADROMOUS)) +
  geom_point(size = 3) +
  scale_color_manual(values = colours) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(lst_percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(lst_percent_var[2], 2), "% variance")) +
  ##geom_text_repel(aes(label = POP), size = 3) +
  theme_minimal()

```

## No Chathams + lst

## Re-calculated eigenvectors for VCF - islands

Read in eigenvector/value data generated by PLINK
```{r}
## read in eigenvectors/values
chalst_vec = read.delim("nochath_nolst.eigenvec", header = TRUE, sep ="\t")
chalst_vall = read.delim("nochath_nolst.eigenval", header = FALSE)

```


Merge metadata and eigenvector data for easy plotting
```{r}
## merge
chalst = merge(chalst_vec, meta, by = "IID")

## turn to df
chalst = as.data.frame(chalst)

```


Calculate percentage of variance explained by each PC for the axes
```{r}
chalst_percent_var = (chalst_vall$V1 / sum(chalst_vall$V1)) * 100

```

## Plot PCA
### PC1/PC2
```{r}
ggplot(chalst, aes(x = PC1, y = PC2, color = POP)) +
  geom_point(size = 3) +
  scale_color_manual(values = colours) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(chalst_percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(chalst_percent_var[2], 2), "% variance")) +
  ##geom_text_repel(aes(label = POP), size = 3) +
  theme_minimal()

```