---
title: "PCA_remake"
author: "Ash Iwikau"
date: "2025-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r message=FALSE}
library("tidyverse")
library("ggplot2")
library("ggrepel")
```

Load in eigenvec/val files
```{r}
eigenvec = read.delim("pca_results.eigenvec", header = TRUE, sep ="\t")
eigenval = read.delim("pca_results.eigenval", header = FALSE)
```

Load in metadata
```{r}
popmetadata = read.table("popmetadatacleaned.txt", header = TRUE, sep = "\t")
```

Merge metadata and eigenvectors
(needs to have a column in common so I just added the IID column to the metadata object)
(pretty sure this only works coz theyre in the same order so be careful in future - find a better way)
```{r}
popmetadata$IID = eigenvec$IID

metadata_eigenvec = merge(eigenvec, popmetadata, by = "IID")
```

Turn into a dataframe (??)
```{r}
dfMetadataEigenvec = as.data.frame(metadata_eigenvec)
```

Calculate percentage of variance explained by each PC for the axes
```{r}
percent_var = (eigenval$V1 / sum(eigenval$V1)) * 100
```

Plot
```{r}
ggplot(dfMetadataEigenvec, aes(x = PC1, y = PC2, color = POPULATION)) +
  geom_point(size = 3) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(percent_var[2], 2), "% variance")) +
  ##geom_text_repel(aes(label = POPULATION), size = 3) +
  theme_minimal()
```

Filter out very divergent pops
```{r}
popfiltered = dfMetadataEigenvec %>% filter(!POPULATION %in% c("Lagoon Saddle Tarn", "Awatotara Stream", "Kawhaki Ck"))
```

Re-plot
```{r}
ggplot(popfiltered, aes(x = PC1, y = PC2, color = POPULATION)) +
  geom_point(size = 3) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(percent_var[2], 2), "% variance")) +
  geom_text_repel(aes(label = POPULATION), size = 3) +
  theme_minimal()
```

It's a bit busy with all the labels but I have yet to figure out of there is a way to have just 1 per cluster 
or if that has to be done manually

## Variations on the PCA

### FULL

```{r}
ggplot(dfMetadataEigenvec, aes(x = PC1, y = PC2, color = MIGRATORY.STATUS)) +
  geom_point(size = 3) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(percent_var[2], 2), "% variance")) +
  ##geom_text_repel(aes(label = POPULATION), size = 3) +
  theme_minimal()
```


```{r}
ggplot(dfMetadataEigenvec, aes(x = PC1, y = PC2, color = ISLAND.MAINLAND)) +
  geom_point(size = 3) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(percent_var[2], 2), "% variance")) +
  ##geom_text_repel(aes(label = POPULATION), size = 3) +
  theme_minimal()
```

### FILTERED

```{r}
ggplot(popfiltered, aes(x = PC1, y = PC2, color = MIGRATORY.STATUS)) +
  geom_point(size = 3) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(percent_var[2], 2), "% variance")) +
  geom_text_repel(aes(label = POPULATION), size = 3) +
  theme_minimal()
```

Left the population labels but maybe change to mig stat?
```{r}
ggplot(popfiltered, aes(x = PC1, y = PC2, color = ISLAND.MAINLAND)) +
  geom_point(size = 3) +
  ##stat_ellipse(type="norm") +
  xlab(paste0("PC1: ", round(percent_var[1], 2), "% variance")) +
  ylab(paste0("PC2: ", round(percent_var[2], 2), "% variance")) +
  geom_text_repel(aes(label = POPULATION), size = 3) +
  theme_minimal()
```

Same as abvoe