---
title: "Polished genome BayeScan processing"
author: "ASH IWIKAU"
date: "2026-01-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages
```{r}
library(ggplot2)
library(tidyverse)
library(stringr)

```

### Load in data
```{r}
## bayescan results
bayescan = read.table("no_islands_test/maf_no_offshore_islands_bayescan_input.g_fst.txt", header = TRUE)

## chromosome positions of snps
pos = read.table("positions_for_bayescan_results.txt", header = FALSE)

## stick together and rename columns
bayescan = cbind(pos , bayescan)
colnames(bayescan)=c("CHROM","POSIT","SNPID","PROB","LOG_PO","Q_VALUE","ALPHA","FST") 

## Rename scaffolds to padded numeric for plotting
bayescan$CHROM <- str_pad(str_extract(bayescan$CHROM, "\\d+"), width = 3, pad = "0", side = "left")


```

#### Assign selection/diversifying/neutral
```{r}
## qval less than 0.01 alpha >=0 = diversifying - everything else neut/bal
bayescan$SELECTION <- ifelse(bayescan$ALPHA>=0&bayescan$Q_VALUE<=0.01,"Diversifying",
                ifelse(bayescan$ALPHA>=0&bayescan$Q_VALUE>0.01,"Neutral","Balancing")) 

## make factor
bayescan$SELECTION<- factor(bayescan$SELECTION)

sum(bayescan$SELECTION=="Diversifying")

```

#### Add small fudge factor to clear 0s
```{r}
## check how many with really small q vals
sum(bayescan$Q_VALUE < 0.0000001)

## add 0.00001 to all values
bayescan$Q = bayescan$Q_VALUE + 0.000001

```

#### Log Q values
```{r}
bayescan$LOG10_Q = -log10(bayescan$Q)

```

#### Make column of significance
```{r}
bayescan$SIG = ifelse(bayescan$Q_VALUE < 0.01, "Signficant", "Non-significant")

bayescan %>% group_by(SIG) %>% tally()
```

#### Write data out to a table
```{r}
#write.table(bayescan, file = "full_bayescan_stats_no_islands.txt", sep = "\t",
#            row.names = FALSE, col.names = TRUE)
```

#### New obj with only significant results
```{r}
sig = bayescan[bayescan$Q_VALUE<=0.01 ,]

write.table(sig, file = "original_bayescan_run_no_islands_significant.txt", sep = "\t",
           row.names = FALSE, col.names = TRUE)
```

## Plotting

#### Set axis names and log cutoffs
```{r}
x_title="-Log10(q-value)" 
y_title=(expression(F[ST]))

log_cutoff = -log10(0.05)
log_cutoff_0.01 = -log10(0.01)
```

#### Plot
```{r}
graph_1 = ggplot(bayescan, aes(x = LOG10_Q, y = FST)) 

```


```{r}
graph_1 + geom_point(aes(fill= SELECTION), pch=21, size=2)+ 
  #geom_text()+ 
  scale_fill_manual(name="Selection",values=c("white","red","orange"))+ 
  labs(x=x_title)+ 
  labs(y=y_title)+ 
  theme(axis.title=element_text(size=12, family="Helvetica",face="bold"), legend.position="none")+ 
  theme(axis.text.x=element_text(colour="black"))+ 
  theme(axis.text.y=element_text(colour="black",size=12))+ 
  theme(axis.text.x=element_text(colour="black",size=12))+ 
  theme(panel.border = element_rect(colour="black", fill=NA, size=3),  
        axis.title=element_text(size=18,colour="black",family="Helvetica",face="bold")) +
  theme_classic() +
  geom_vline(xintercept = log_cutoff, linetype="dashed", color="red", linewidth=1) +
  geom_vline(xintercept = log_cutoff_0.01, linetype="dashed", color="blue", linewidth=1)

```


```{r}
graph_1 + 
  geom_point(aes(fill = ifelse(SELECTION == "Diversifying", "Significant", "Background")), pch=21, size=2) + 
  #geom_text()+ 
  scale_fill_manual(name = "Selection", values = c("Significant" = "red", "Background" = "transparent")) + 
  labs(x=x_title)+ 
  labs(y=y_title)+ 
  theme(axis.title=element_text(size=12, family="Helvetica",face="bold"), legend.position="none")+
  theme(axis.text.x=element_text(colour="black"))+ 
  theme(axis.text.y=element_text(colour="black",size=12))+ 
  theme(axis.text.x=element_text(colour="black",size=12))+ 
  theme(panel.border = element_rect(colour="black", fill=NA, size=3),  
        axis.title=element_text(size=18,colour="black",family="Helvetica",face="bold")) +
  theme_classic() +
  geom_vline(xintercept = log_cutoff, linetype="dashed", color="red", linewidth=1) +
  geom_vline(xintercept = log_cutoff_0.01, linetype="dashed", color="blue", linewidth=1) 

```

## Manhattans

#### Check how many are significant per chrom
```{r}
sig %>% group_by(CHROM) %>% tally()

```

43/71 are on scaffold 17! Yay

#### Sort by FST value
```{r}
sorted = sig[order(sig$FST, decreasing = TRUE),]

```

#### Make df for ease
```{r}
df = as.data.frame(bayescan)

```

#### Cumulative positions for plotting
```{r}
## Calculate
cumulative_position = df  %>%
  group_by(CHROM) %>%
  summarise(max_bp = max(POSIT)) %>%
  mutate(bp_add = lag(cumsum(max_bp), default = 0)) %>%
  select(CHROM, bp_add)

## add back
df <- df %>%
  inner_join(cumulative_position, by = "CHROM") %>%
  mutate(CUMULATIVEBP = POSIT + bp_add)

## set axes in the moddile of each chrom for plotting
axis_set = df |>
  group_by(CHROM) |>
  summarize(center = mean(CUMULATIVEBP))


```

#### Plot full manhattan
```{r}
manhattan_fst = ggplot(df, 
                         aes(x = CUMULATIVEBP, y = FST, color = as_factor(CHROM))) + 
  geom_point(size = 3 , alpha = 0.5, ) +
  scale_x_continuous( label = axis_set$CHROM, breaks = axis_set$center,) +
  scale_color_manual(values = rep(c("darkgrey", "#363636"), unique(length(axis_set$CHROM)) )) +
  geom_point(data=subset(df, SIG=="Signficant"), color="red", size=3, alpha = 0.5) +
  labs( x = NULL, y = (expression(F[ST])) , ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.text.x = element_text(angle = 90, size = 6, vjust = 0.5))

manhattan_fst

```

#### Plot outlier scaffold manhattan

```{r}
scaff17 = ggplot(df %>% filter(CHROM == "017"), 
                       aes(x = POSIT, y = FST, color = as_factor(CHROM))) + 
  geom_point(size = 3, alpha = 0.5, ) +
  scale_x_continuous() + 
  scale_color_manual(values = rep(c("darkgrey", "#363636"), unique(length(axis_set$CHROM)) )) +
  geom_point(data=subset(df %>% filter(CHROM == "017"), SIG=="Signficant"), color="red", size=3, alpha = 0.5,) +
  labs( x = "Scaffold 17", y = (expression(F[ST])) , ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.text.x = element_blank())

scaff17
```



